### 2018-01-23 latex/egsr17-cgf/main-body.tex
started working on supplemental document with extended derivations

\title[Draft]%
{Evaluation of Higher Order Diffusion against Variable Eddington Factor Methods for Rendering Participating Media}

@@

% uncomment for using teaser
% \teaser{
%  \includegraphics[width=\linewidth]{eg_new}
%  \centering
%   \caption{New EG Logo}
% \label{fig:teaser}
%}

@@

\begin{abstract}

Classical diffusion approximation (CDA) is very competitive for rendering highly scattering participating media, because in these scenarios, Monte Carlo based methods suffer from high computational cost. Variable Eddington Factor methods, such as flux-limited diffusion (FLD), have shown to further improve the results from CDA. In this paper, we look at CDA as a degenerate case of the $P_1$-method and investigate the open question, if going higher order gives a viable alternative to FLD. We derive the time-independent real-valued $P_N$-equations and introduce a framework for generating finite difference stencils directly from the computer algebra representation of the $P_N$-equations for arbitrary order. We further present a staggered-grid $P_N$-solver and compare it against FLD for various standard problems.

@@

Light transport in participating media is governed by the radiative transfer equation(RTE), which was first introduced in astrophysics by Chandrashekar~\cite{Chandrasekhar60} and later used for rendering in graphics by Kajiya~\cite{Kajiya86}. Today, this equation is typically solved using Monte Carlo based methods. However, in strongly scattering media or in the presence of highly anisotropic phase functions, these methods can become unfeasible, due to high computational cost which comes from tracing long particle paths with many scattering events.

@@

CDA suffers from severe energy loss close to regions with strong density gradients. This problem is addressed by the variable eddington factor(VEF)-formalism, among which flux-limited diffusion (FLD) is the most prominent. FLD was first introduced in astrophysics by~\cite{Pomraining78} and was introduced to graphics by Koerner et al.~\cite{Koerner14}. Unfortunately, both CDA and FLD do not support anisotropic phase functions.

The idea behind VEF is to replace the diffusion coefficient in the diffusion equation of CDA by a non-linear diffusion coefficient. This diffusion coefficient is being derived by seeing it as a mean to interpolate between the diffusive and pure transport regime. The diffusive regime is where thick highly scattering media is present and causes photons to change directions in short succession, while the transport regime in thin and highly absorbing media causes photons to travel long straight lines. In the absence of scattering (e.g. pure vacuum), the non-linear diffusion coefficient turns the diffusion equation into an advection equation.

@@

The $P_N$-equations are derived from the radiative transfer equation~(RTE), which expresses the change of the radiance field $L$, with respect to an infinitesimal change of position into direction $\omega$ at point $\vec{x}$:
\begin{align}
%\label{eq:rte}
\left(\nabla\cdot\omega\right)L\left(\vec{x}, \omega \right)
=&
-\sigma_t\left(\vec{x}\right) L\left(\vec{x}, \omega \right)\\
&
+\sigma_s\left(\vec{x}\right) \int_{\Omega}
{
p\left(\omega'\cdot\omega\right)L\left(\vec{x}, \omega \right)\ud\omega'
}\\
&
+Q\left(\vec{x}, \omega\right)
\end{align}

The LHS is the transport term and we refer to the terms on the RHS as collision, scattering and source term respectively.

@@

@@

First, angular dependent quantities in the RTE, such as the radiance field $L$ (or the emission field $Q$) are projected into SH coefficients:
\begin{align}
L^{l,m}\left(\vec{x}\right)
=
\int_\Omega{L\left(\vec{x}, \omega\right)\SHBR^{l,m}\ud\omega}
\end{align}

The reconstruction $\hat{L}$ will have an error due to truncation at order $N$:
\begin{align}
L\left(\vec{x}, \omega\right)
\approx
\hat{L}\left(\vec{x}, \omega\right) =
\sum_{l=0}^{N}
{
\sum_{m=-l}^{l}
{
L^{l,m}\left(\vec{x}\right)\SHBR^{l,m}\left(\omega\right)
}
}
=
\sum_{l,m}
{
L^{l,m}\left(\vec{x}\right)\SHBR^{l,m}\left(\omega\right)
}
\end{align}

The angular dependent quantities are replaced with their SH reconstruction in the RTE. The $P_N$-equations are found by multiplying each term with the conjugate complex of the real-valued SH basis function $\overline{\SHBR^{l,m}}$, followed by integration over solid angle. In order to arrive at a form, which does not depend on angle anymore, a set of transformations and identities are applied. Of particular importance are the orthogonality property
\begin{align}
\int_{\Omega}{ \SHBC^{l,m}\overline{\SHBC^{l',m'}}\ud\omega} = \delta_{l,l'}\delta_{m,m'}
\end{align}

and the following recursion relation
\begin{align}
\omega_x\overline{\SHBC^{l, m}}=
&
-\frac{1}{2}c^{{l-1,m-1}}\overline{Y^{l-1, m-1}}+\frac{1}{2}d^{{l+1,m-1}}\overline{Y^{l+1, m-1}}\\
&
+\frac{1}{2}e^{{l-1,m+1}}\overline{Y^{l-1, m+1}}-\frac{1}{2}f^{{l+1,m+1}}\overline{Y^{l+1, m+1}}\\
\omega_y\SHBC^{l, m}=
&
\frac{i}{2}c^{{l-1,m-1}}\overline{Y^{l-1, m-1}}-\frac{i}{2}d^{{l+1,m-1}}\overline{Y^{l+1, m-1}}\\
&
+\frac{i}{2}e^{{l-1,m+1}}\overline{Y^{l-1, m+1}}-\frac{i}{2}f^{{l+1,m+1}}\overline{Y^{l+1, m+1}}\\
\omega_y\SHBC^{l, m}=
&
a^{{l-1,m}}\overline{Y^{l-1, m}}+b^{{l+1,m}}\overline{Y^{l+1, m}}
\end{align}

with 
\begin{align}
&a^{l,m}= \sqrt{\frac{\left(l-m+1\right)\left(l+m+1\right)}{\left(2l+1\right)\left(2l-1\right)}},
b^{l,m}= \sqrt{\frac{\left(l-m\right)\left(l+m\right)}{\left(2l+1\right)\left(2l-1\right)}}\\
&c^{l,m}= \sqrt{\frac{\left(l+m+1\right)\left(l+m+2\right)}{\left(2l+3\right)\left(2l+1\right)}},
d^{l,m}= \sqrt{\frac{\left(l-m\right)\left(l-m-1\right)}{\left(2l+1\right)\left(2l-1\right)}}\\
&e^{l,m}= \sqrt{\frac{\left(l-m+1\right)\left(l-m+2\right)}{\left(2l+3\right)\left(2l+1\right)}},
f^{l,m}= \sqrt{\frac{\left(l+m\right)\left(l+m-1\right)}{\left(2l+1\right)\left(2l-1\right)}}
\end{align}

@@

Note that the real-valued SH basis functions are different, depending on $m$, which means that we get three different versions of the $P_N$-equations (for $m<0$, $m=0$ and $m>0$). Further note, that we need to expand the real-valued SH basis functions in order to be able to apply the given identities above. This causes the equations to become very large and difficult to deal with when applying transformations and identities.

It becomes even more of a problem later, when we want to expand the spatially discretized equations for a particular choice of truncation order $N$. Manually creating a FD stencil from hundreds of terms, including complex numbers and derivatives, is very challenging for first order and becomes impossible for higher order.

@@

In order to tackle this problem, we created a system for representing algebraic expressions as an expression tree in Python. Our framework can represent terms, integrals, derivatives, sums, variables, functions and special symbols, such as the Kronecker delta or imaginary number. It supports manipulations, such as the application of distributive law, substitution, rearranging nested integrals and application of the identities given above.

@@

<show some codesnippet for generating expressions and applying manipulations?>

@@

@@

As we will see later in section~\ref{sec:results}, our solver needs to support staggered grids in order to avoid severe oscillation artefacts which arise, when all coefficients are located at the voxel center. With staggered grids, coefficients can live on the intersections of a subgrid with double resolution. So instead of voxel centers alone, coefficients can live on the center of voxel faces, voxel edges or the intersection between voxels.

@@

@@

Additional active coefficients (which contribute additional rows and columns in $A$ and $\vec{Q}$) are required to respect the boundary correctly (see Figure~\ref{fig:staggeredgrid}). This requires careful managegment of additional coefficients in boundary voxels, which is done transparently by the C++ framework.

\todo[inline]{The original todo note withouth changed colours.\newline Here's another line.}

\begin{figure}[h]
\centering
\missingfigure{\Huge Nice this triangle?}
\caption{Some caption}
\label{nicefigure}
\end{figure}

<picture showing staggered boundary coefficients>
%\label{fig:staggeredgrid}

%Matrix $A$ and vector $\vec{Q}$ are constructed after applying the stencil for every voxel.
The generated linear system becomes very large for high order and dense FD grids. The number of rows is determined by the number of voxels times the number of SH coefficients per voxel. The matrix $A$ is squared and fortunately very sparse. It further is non-symmetric and also not diagonal dominant, which rules out many standard methods for solving linear systems. We adress this by solving the normal form $A^TA\vec{u}=A^T\vec{Q}$ instead. This gives a symmetric and positive definit system matrix $A^TA$ (although with a higher condition number).

@@

\subsection{Pointsource problem}
\subsection{Nebulae/Cloud}

>>>>

\title[Draft]%
{
%Evaluation of Higher Order Diffusion against Variable Eddington Factor Methods for Rendering Participating Media
A Deterministic Method for Simulation of Light Transport in Participating Media based on Higher Order Spherical Harmonics
}

@@

% uncomment for using teaser
\teaser
{
\centering
\begin{subfigure}{0.2\linewidth}
\missingfigure{}
\caption{$CDA/P_1$}
\end{subfigure}%
\hspace{0.05\linewidth}
\begin{subfigure}{0.2\linewidth}
\missingfigure{}
\caption{FLD}
\end{subfigure}%
\hspace{0.05\linewidth}
\begin{subfigure}{0.2\linewidth}
\missingfigure{}
\caption{$P_5$}
\end{subfigure}%
\hspace{0.05\linewidth}
\begin{subfigure}{0.2\linewidth}
\missingfigure{}
\caption{Groundtruth}
\end{subfigure}%
%\includegraphics[width=\linewidth]{figures/eg_new}
\icaption
{
Flux-limited diffusion (b) is an ad-hoc extension to the classical diffusion approximation (a). It can be considered the state of the art among deterministic methods for simulating light transport in highly scattering participating media, which is a common challenge for brute-force path tracing (d). We add the $P_N$-method (c) to the arsenal of available deterministic methods in rendering and investigate its benefits and tradeoffs against flux-limited diffusion.
}
\label{fig:teaser}
}

@@

\begin{abstract}
%Classical diffusion approximation (CDA) is very competitive for rendering highly scattering participating media, because in these scenarios, Monte Carlo based methods suffer from high computational cost. Variable Eddington Factor methods, such as flux-limited diffusion (FLD), have shown to further improve the results from CDA. In this paper, we look at CDA as a degenerate case of the $P_1$-method and investigate the open question, if going higher order gives a viable alternative to FLD. We derive the time-independent real-valued $P_N$-equations and introduce a framework for generating finite difference stencils directly from the computer algebra representation of the $P_N$-equations for arbitrary order. We further present a staggered-grid $P_N$-solver and compare it against FLD for various standard problems.
Rendering highly scattering participating media is a common challenge for brute-force path-tracing. Classical diffusion approximation (CDA) is a competitive alternative, as it reduces the problem to solving a simple diffusion equation. Variable eddington factor methods, such as flux-limited diffusion (FLD), improve CDA by making the diffusion coefficient non-linear. Both methods are based on the spherical harmonics expansion of the radiance field, truncated after the first order. In this paper, we investigate the open question, if going higher order provides a viable alternative to the ad-hoc fixes to diffusion by FLD. Increasing the order introduces a growing set of increasingly complex coupled partial differential equations, called $P_N$-equations, which become very difficult to work with for higher order. We introduce a computer algebra framework for representing and manipulating mathematical equations and use it to derive the time-independent real-valued $P_N$-equations at arbitrary order. We further present a staggered-grid $P_N$-solver and generate its stencil code directly from the expression tree of the $P_N$-equations. Both, the computer algebra framework, as well as the solver are made avaiable open source. Finally, we discuss how our method compares against FLD for various standard problems.

@@

Light transport in participating media is governed by the radiative transfer equation~(RTE), which was first introduced in astrophysics by Chandrashekar~\cite{Chandrasekhar60} and later used for rendering in graphics by Kajiya~\cite{Kajiya86}. Today, this equation is typically solved using Monte Carlo based methods. However, in strongly scattering media or in the presence of highly anisotropic phase functions, these methods can become unfeasible, due to the high computational cost which comes from tracing long particle paths with many scattering events.

@@

CDA suffers from severe energy loss close to regions with strong density gradients. This problem is addressed by the variable eddington factor(VEF)-formalism, among which flux-limited diffusion (FLD) is the most prominent. FLD was first introduced in astrophysics by~\cite{Levermore81} and was introduced to graphics by Koerner et al.~\cite{Koerner14}.

The idea behind VEF is to replace the diffusion coefficient in the diffusion equation of CDA by a non-linear diffusion coefficient. This new diffusion coefficient is being derived by seeing it as a mean to spatially blend between the diffusive and pure transport regime. The diffusive regime is where thick highly scattering media is present and causes photons to change directions in short succession, while the transport regime in thin and highly absorbing media causes photons to travel long straight lines. In the absence of scattering (e.g. pure vacuum), the non-linear diffusion coefficient turns the diffusion equation into an advection equation.

@@

In this section we present how to optain the real-valued $P_N$-equations. For brevity and readability, we only give a brief outline here and refer to the supplemental material for the detailed derivation.

The $P_N$-equations are derived from the radiative transfer equation, which expresses the change of the radiance field $L$, with respect to an infinitesimal change of position into direction $\omega$ at point $\vec{x}$:
\begin{align}
%\label{eq:rte}
\left(\nabla\cdot\omega\right)L\left(\vec{x}, \omega \right)
=&
-\sigma_t\left(\vec{x}\right) L\left(\vec{x}, \omega \right)\nonumber\\
&
+\sigma_s\left(\vec{x}\right) \int_{\Omega}
{
p\left(\omega'\cdot\omega\right)L\left(\vec{x}, \omega \right)\ud\omega'
}\nonumber\\
&
+Q\left(\vec{x}, \omega\right)\nonumber
\end{align}

The LHS is the transport term and we refer to the terms on the RHS as collision, scattering and source term respectively. The symbols $\sigma_t$, $\sigma_s$, $p$ and $Q$ refer to the extinction coefficient, scattering coefficient, phase function and emission term respectively.

@@

\begin{figure}[h]
\centering
\missingfigure{test}
\icaption{Structure of coefficient matrix $A$ and solution vector $\vec{u}$ after discretization of the $P_N$-equations on a finite difference grid.}
\end{figure}

@@

First, the angular dependent quantities in the RTE, such as the radiance field $L$ (or the emission field $Q$) are projected into SH coefficients:
\begin{align}
L^{l,m}\left(\vec{x}\right)
=
\int_\Omega{L\left(\vec{x}, \omega\right)\SHBR^{l,m}\ud\omega}
\end{align}

The reconstruction $\hat{L}$ has an error due to truncation at order $N$:
\begin{align}
L\left(\vec{x}, \omega\right)
\approx
\hat{L}\left(\vec{x}, \omega\right) =
\sum_{l=0}^{N}
{
\sum_{m=-l}^{l}
{
L^{l,m}\left(\vec{x}\right)\SHBR^{l,m}\left(\omega\right)
}
}
=
\sum_{l,m}
{
L^{l,m}\left(\vec{x}\right)\SHBR^{l,m}\left(\omega\right)
}
\end{align}

The angular dependent quantities are replaced with their SH reconstruction in the RTE. The $P_N$-equations are found by multiplying each term with the conjugate complex of the real-valued SH basis function, followed by integration over solid angle. In order to arrive at a form, which does not depend on angle anymore, a set of transformations and identities are applied. Of particular importance are the orthogonality property
\begin{align}
\int_{\Omega}{ \SHBC^{l,m}\overline{\SHBC^{l',m'}}\ud\omega} = \delta_{l,l'}\delta_{m,m'}
\end{align}

and recursion relations such as the following:
\begin{align}
\omega_z\SHBC^{l, m}=
&
a^{{l-1,m}}\overline{Y^{l-1, m}}+b^{{l+1,m}}\overline{Y^{l+1, m}}
\end{align}

with 
\begin{align}
&a^{l,m}= \sqrt{\frac{\left(l-m+1\right)\left(l+m+1\right)}{\left(2l+1\right)\left(2l-1\right)}},
b^{l,m}= \sqrt{\frac{\left(l-m\right)\left(l+m\right)}{\left(2l+1\right)\left(2l-1\right)}}
\end{align}

@@

Note that the real-valued SH basis functions are different, depending on $m$, which means that we get three different versions of the $P_N$-equations (for $m<0$, $m=0$ and $m>0$). Further note, that we need to expand the real-valued SH basis functions into their complex-valued definition, in order to be able to apply the given identities above. This causes the equations to become large and difficult to deal with when applying transformations and identities.

It becomes even more of a problem later, when we want to expand the spatially discretized equations for a particular choice of truncation order $N$. Manually creating a FD stencil from so many terms is challenging for first order and becomes increasingly difficult and error prone for higher order.

@@

\begin{figure}[h]
\centering
\begin{subfigure}{0.45\columnwidth}
\missingfigure{test}
\end{subfigure}%
\hspace{0.05\columnwidth}
\begin{subfigure}{0.45\columnwidth}
\missingfigure{test2}
\end{subfigure}%

\begin{subfigure}{0.45\columnwidth}
\missingfigure{test}
\end{subfigure}%
\hspace{0.05\columnwidth}
\begin{subfigure}{0.45\columnwidth}
\missingfigure{test2}
\end{subfigure}%

\icaption{The computer algebra representation is constructed using API commands provided by our framework (top left) and represented internally as an expression tree (top right). Additional API commands (bottem left) allow valid transformations, which modify the expression tree accordingly (bottom right).}
\end{figure}

In order to tackle this problem, we created a system for representing algebraic expressions as an expression tree. Our framework can represent terms, integrals, derivatives, sums, variables, functions and special symbols, such as the Kronecker delta or imaginary number. It supports manipulations, such as the application of distributive law, substitution, rearranging nested integrals and application of the identities given above.

@@

@@

\begin{figure*}[t]
\centering
\missingfigure{$P_N$-solver overview: generate stencil code $\rightarrow$ build system $\rightarrow$ solve $\rightarrow$ render}
\icaption{Overview of our $P_N$-solver. After generating the stencil source code from the expression trees representing the $P_N$-equations, the linear system $A\vec{u}=\vec{Q}$ is built using RTE parameter fields and additional user input, such as grid resolution and type of boundary conditions. The resulting system is solved for $\vec{u}$, which is then used in our rendering application.}
\label{fig:pnsolver}
\end{figure*}

@@

As we see in figure~\ref{fig:artefacts}, our solver needs to support staggered grids in order to avoid severe oscillation artefacts which arise, when all coefficients are located at the voxel center. With staggered grids, coefficients can live on the intersections of a subgrid with double resolution. So instead of voxel centers alone, coefficients can live on the center of voxel faces, voxel edges or the intersection between voxels.

@@

\begin{figure}[h]
\centering
\begin{subfigure}{0.45\columnwidth}
\missingfigure{test}
\end{subfigure}%
\hspace{0.05\columnwidth}
\begin{subfigure}{0.45\columnwidth}
\missingfigure{test2}
\end{subfigure}%
\icaption{The solution of our solver for the 2D-checkerboard problem with collocated grids (left), where all coefficients are located at the cell centers. It suffers from severe oscillating artefacts, which are removed by offseting the coefficient locations using staggered grids (right).}
\label{fig:artefacts}
\end{figure}

@@

Additional active coefficients (which contribute additional rows and columns in $A$ and $\vec{Q}$) are required to respect the boundary correctly (see Figure~\ref{fig:staggeredgrid}). This requires careful managegment of additional coefficients at boundary voxels, which is done transparently by the C++ framework.

%\todo[inline]{The original todo note withouth changed colours.\newline Here's another line.}

\begin{figure}[h]
\centering
\begin{subfigure}{0.45\columnwidth}
\missingfigure{test}
\end{subfigure}%
\hspace{0.05\columnwidth}
\begin{subfigure}{0.45\columnwidth}
\missingfigure{test2}
\end{subfigure}%
\icaption{Staggered grids cause a shifted boundary on the right and upper side of the domain (left). The boundary is correct after introducing additional unknowns (red) at boundary voxels (right).}
\label{fig:staggeredgrid}
\end{figure}

%Matrix $A$ and vector $\vec{Q}$ are constructed after applying the stencil for every voxel.
The generated linear system becomes very large for high order and resolution. The number of rows is determined by the number of voxels times the number of SH coefficients per voxel. The matrix $A$ is squared and fortunately very sparse. It further is non-symmetric and also not diagonal dominant, which rules out many standard methods for solving linear systems. We adress this by solving the normal form $A^TA\vec{u}=A^T\vec{Q}$ instead. This gives a symmetric and positive definit system matrix $A^TA$ (although with a higher condition number).

@@

\subsection{Rendering}

\begin{itemize}
	\item explain that we use single scattered light as emission field
	\item explain our single scattering path tracer using the $P_N$-solution as indirect light
\end{itemize}

\subsection{2D checkerboard}
\subsection{Pointsource problem}
\subsection{Nebulae/Cloud}
